---
title: "Dataframes - In Class lab"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")


head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}
glimpse(grants)

```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
grants$org <- grants$organization 
grants

```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
is.vector(grants$organization)

```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
mean(grants$total_amount)
```


- What was the dollar amount of the largest grant?

```{r}
max(grants$total_amount)
```


- What was the dollar amount of the smallest grant?

```{r}
min(grants$total_amount)
```


- Which organization received the largest grant?

```{r}
max(grants$organization)

```


- Which organization received the smallest grant?

```{r}
min(grants$organization)
```

- How many grants were awarded in 2010?

```{r}
sum(grants$start_year == 2010)
```

- How much was awarded in 2010? 

```{r}
sum(grants$total_amount[grants$start_year == 2010])

sum(grants[grants$start_year == 2010, 'total_amount'])
```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
expend_df <- data.frame(USPersonalExpenditure)
expend_df
```


What are the current column names of your dataframe?
```{r}
colnames(expend_df)
```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
rownames(expend_df)
```


Add a column "Category" to your data frame that contains the rownames

```{r}
expend_df$Category <- rownames(expend_df)

expend_df
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
tidy_expend <- pivot_longer(
  data = expend_df,    
  cols = c("X1940", "X1945", "X1950", "X1955", "X1960"),  
  names_to = "Year",            
  values_to = "Expenditure"     
)

tidy_expend
```


How much money was spent on personal care in 1940?
```{r}
expend_df["Personal Care", "X1940"]
```


How much money was spent on Food and Tobacco in 1960?
```{r}
expend_df["Food and Tobacco", "X1960"]

```


What was the highest expenditure category in 1960?
```{r}
expend_df$Category[which.max(expend_df$X1960)]

```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}
lowest_category <- function(year) {
  expend_df$Category[which.min(expend_df[[year]])]
}

lowest_category("X1960")
```

Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}
sapply(c("1960","1970","1980","1990"), lowest_category)

```
